---
title: "Preliminary analysis of flash and mash result"
# author: Yanyu Liang
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

# Set up

```{r setup}
library(ggplot2)
library(mashr)
library(biomaRt)
library(dplyr)
library(topGO)
library(pander)
library(rmeta)
library(reshape2)
panderOptions('knitr.auto.asis', FALSE)
panderOptions('table.split.table', Inf)
```

# Read data

```{r}
flash <- readRDS('../output/gtex_flash.rds')
mash <- readRDS('../output/gtex_mash.rds')
eqtl <- read.table('../output/strong_eqtl_names.txt')$V1
## get gtex color
gtex.color <- read.table(url('https://github.com/stephenslab/gtexresults/raw/master/data/GTExColors.txt'), sep = '\t', comment.char = '')
gtex.color <- gtex.color[c(1:6,9:18,21:23,26:30,32,33,35,36,38:53), 1:2]
tissue.color <- as.character(gtex.color[,2])
ntissue <- nrow(gtex.color)
```

# Factor 5

See details of factor 5 at [here](run_flashr.html)

```{r}
l <- flash$EL
dim(l)
l.five <- l[, 5]
top.tissue.idx <- order(abs(flash$EF[, 5]), decreasing = T)[1 : 3]
hist(l.five)
top.ten.idx <- order(l.five, decreasing = T)[1:10]
bottom.ten.idx <- order(l.five, decreasing = F)[1:10]
```

## Ten eQTL with highest loading

```{r, fig.height = 15, fig.width = 8}
par(mfrow = c(5, 2))
for(i in 1 : 10) {
  mash_plot_meta(mash$posterior, top.ten.idx[i], color = meta.colors(lines = tissue.color), main = paste0(i, ': ', eqtl[i]))
}
```

## Ten eQTL with lowest loading

```{r, fig.height = 15, fig.width = 8}
par(mfrow = c(5, 2))
for(i in 1 : 10) {
  mash_plot_meta(mash$posterior, bottom.ten.idx[i], color = meta.colors(lines = tissue.color), main = paste0(i, ': ', eqtl[i]))
}
```

# Gene ontology enrichment analysis

EQTLs with top and bottom loadings (50 respectively) are used as positive set and 500 eQTLs with most modest loading as background set.

```{r, results='asis'}
annotateEQTL <- function(eqtl, idx, ensembl) {
  ensg <- data.frame(t(sapply(eqtl[idx], function(str) {
    temp <- strsplit(as.character(str), '_')[[1]]
    x <- temp[1]
    chr <- temp[2]
    s <- temp[3]
    return(c(strsplit(x, '.', fixed = T)[[1]][1], chr, s))
  })))
  colnames(ensg) <- c('ensembl_gene_id', 'eqtl_chr', 'eqtl_start')
  ensg$idx <- idx
  ensg$ensembl_gene_id <- as.character(ensg$ensembl_gene_id)
  re <- getBM(
    attributes = c('ensembl_gene_id', 'hgnc_symbol', 'chromosome_name', 'start_position', 'end_position', 'strand', 'go_id', 'name_1006', 'uniprot_genename', 'uniprot_swissprot_accession'), 
    filters = 'ensembl_gene_id',
    values = ensg[, 1],
    mart = ensembl)
  re <- inner_join(re, ensg, by = 'ensembl_gene_id')
  return(re)
}
genGene2GO <- function(g2g, re) {
  gene <- unique(re$ensembl_gene_id)
  for(x in gene) {
    g2g[[x]] <- re[re$ensembl_gene_id == x, 'go_id']
  }
  return(g2g)
}
ensembl <- useMart(biomart="ENSEMBL_MART_ENSEMBL", host="feb2014.archive.ensembl.org", path="/biomart/martservice", dataset="hsapiens_gene_ensembl")
positive.idx <- c(order(l.five, decreasing = T)[1:50], order(l.five, decreasing = F)[1:50])
control.idx <- order(abs(l.five), decreasing = F)[1:500]
positive.re <- annotateEQTL(eqtl, positive.idx, ensembl)
control.re <- annotateEQTL(eqtl, control.idx, ensembl)
gene2go <- list()
gene2go <- genGene2GO(gene2go, positive.re)
gene2go <- genGene2GO(gene2go, control.re)
all.gene <- union(unique(positive.re$ensembl_gene_id), unique(control.re$ensembl_gene_id))
gene.list <- factor(as.integer(all.gene %in% unique(positive.re$ensembl_gene_id)))
names(gene.list) <- all.gene
map <- list(BP = 'Biological process', MF = 'Molecular function', CC = 'Cellular component')
for(i in c('BP', 'MF', 'CC')) {
  cat('##', map[[i]])
  cat('\n')
  GOdata <- new("topGOdata", ontology = i, allGenes = gene.list, annot = annFUN.gene2GO, gene2GO = gene2go)
  resultFisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")
  allRes <- GenTable(GOdata, classicFisher = resultFisher)
  pander(allRes, caption = i)
  cat('\n')
  cat('\n')
}

```

It turns out that these top genes are enriched in cellular respiration process. 

# Overview of selected gene sets

```{r}
df <- data.frame(loading = l.five[c(positive.idx, control.idx)], label = c(rep('positive', length(positive.idx)), rep('control', length(control.idx))))
ggplot(df) + geom_histogram(aes(x = loading, group = label, fill = label), binwidth = 1) + ggtitle('Histogram of loading')
```

## Posterior mean comparison across top three tissues

```{r}
df.load <- mash$posterior$result$PosteriorMean[positive.idx, top.tissue.idx]
pairs(df.load, main = 'Pairwise comparison of top three tissues in positive set')
df.load <- mash$posterior$result$PosteriorMean[control.idx, top.tissue.idx]
pairs(df.load, main = 'Pairwise comparison of top three tissues in control set')
```

So, it seems that the ones with strong loading on factor 5 are driven by tissue specificity but not strong correlation between these top tissues. 

## Posterior effect size of control eQTLs

```{r, fig.height = 15, fig.width = 8}
random.idx <- sample(control.idx, 10, replace = F) 
par(mfrow = c(5, 2))
for(i in 1 : 10) {
  mash_plot_meta(mash$posterior, random.idx[i], color = meta.colors(lines = tissue.color), main = paste0(i, ': ', eqtl[i]))
}
```

# Cellular respiration (CR)

GO term 'GO:0055114' is used (represent oxidation-reduction process)

## Loading

Contrast eQTL in cellular respiration with the ones unrelated to cellular respiration on the loading of factor 5. 

```{r}
re.go <- annotateEQTL(eqtl, 1 : length(eqtl), ensembl)
cr.gene <- re.go[re.go$go_id == 'GO:0055114', 'idx']
label <- rep('Not CR', length(l.five))
label[unique(cr.gene)] <- 'CR'
df <- data.frame(loading = l.five, label = label)
ggplot(df, aes(x = abs(loading), color = label)) + stat_ecdf()
ks.test(abs(df$loading[df$label == 'Not CR']), abs(df$loading[df$label == 'CR']), alternative = 'g')
wilcox.test(abs(df$loading[df$label == 'Not CR']), abs(df$loading[df$label == 'CR']), alternative = 'l')
```

The result supports the hypothesis that the eQTLs with strong loading on factor 5 (most likely to be tissue-specific eQTLs in top tissues, *i.e.* heart and skeletal muscle) are enriched in celluar respiration process. It makes sense since muscle tissue needs huge amount of ATP which needs highly active celluar repiration.

## Effect size

Contrast effect size 

```{r}
df <- data.frame(mash$posterior$result$PosteriorMean[, ])
df.se <- data.frame(mash$posterior$result$PosteriorSD[, ])
col <- colnames(df)
df.sig <- data.frame(abs(df) > df.se * 1.96)
df.sig$label <- label
df.sig <- melt(df.sig, id.vars = 'label')
ggplot(df.sig) +
  geom_bar(aes(x = variable, group = value, color = value), position = "fill") +
  facet_grid(label~.) +
  scale_fill_manual(values = tissue.color, guide = FALSE) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, color = tissue.color)) +
  ggtitle('Proportion of non-zero effect gene')
df.prop <- df.sig %>% group_by(variable, label) %>%
  summarise(prop = sum(value) / length(value))
df <- data.frame(CR = df.prop$prop[df.prop$label == 'CR'], NCR = df.prop$prop[df.prop$label == 'Not CR'], tissue = df.prop$variable[df.prop$label == 'CR'])
ggplot(df) + geom_point(aes(x = NCR, y = CR, color = tissue)) +
  scale_color_manual(values = tissue.color, guide = FALSE) + geom_abline(slope = 1, intercept = 0) +
  ggtitle('Proportion of non-zero effect gene (CR vs not CR)')
  
```

Overall, genes involved in cellular respiration is more likely to have non-zero posterior effect (95% credible interval).

# Save annotated eQTLs

```{r}
gz1 <- gzfile("../output/strong_eqtl_annotated.txt.gz", "w")
write.table(re.go, gz1, row.names = F, col.names = T)
close(gz1)
```